Uncovering underlying musical content in the temporal domain \\ \small Leveraging deep learning, inductive bias, and music aural skills to learn embedding spaces from raw waveforms with applications to MIR downstream tasks

### 2023-04-07 MasterThesis.tex
Updates from Overleaf

\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb, latexsym}
\usepackage{sidecap}
%---tikz---
\usepackage{tikz}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{fadings}

@@

%Title parameters
\title{%
  Uncovering underlying musical content in the temporal domain \\
  \small Leveraging deep learning, inductive bias, and aural skills to learn embedding spaces from raw waveforms with applications to MIR downstream tasks}

@@

%Introduction of the project
\input{introduction/introduction}
%Task definition
%%\input{task definition}
%Methods
\input{methods/methods}
%Results
\input{results/results}
%Discussion
\input{discussion/discussion}

@@

% appendices come here
\bibliographystyle{naturemag}
\bibliography{bibliography}

\input{appendix/appendix}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb, latexsym}
\usepackage{sidecap}

%---tikz---

\usepackage{tikz}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{fadings}
\usetikzlibrary{positioning}

@@

%Title parameters
\title{%
  Uncovering underlying musical content in the temporal domain \\
  \small Leveraging deep learning, inductive bias, and music aural skills to learn embedding spaces from raw waveforms with applications to MIR downstream tasks}

@@

%Introduction of the project
\input{introduction/introduction}
%Task definition
%%\input{task definition}
%Methods
\input{methods/methods}
%Results
\input{results/results}
%Discussion
\input{discussion/discussion}
%Future work
\input{future work/future work}
\newpage

@@

% appendices come here
\bibliographystyle{naturemag}
\bibliography{bibliography/references.bib}

\newpage
\input{appendix/appendix}

### 2023-04-07 introduction/Subsections/Assumptions.tex
Updates from Overleaf

This thesis assumes that high-level musical concepts are agnostic and invariant to the unique tonal quality of the waveform, much like how a sheet music composition can be performed in infinite ways using different instruments and styles while retaining its musical essence.

We rely on a flexible and adaptable blueprint for musical expression, allowing performers to interpret and express the same musical concepts differently creatively.

Different waveforms can express the same musical concepts despite having distinct timbres or production styles. Still, it is crucial to distinguish between low-level and high-level musical features when analyzing or comparing different musical pieces. High-level features, such as melody, harmony, rhythm, tempo, form, and expression, are more abstract and can be recognized regardless of the specific waveform or production style.

The thesis relies on the objective and comprehensive perspective on a piece's musical content and meaning, much like reading sheet music can reveal the underlying musical structure and intention behind a piece of music.

One sheet music composition can be performed and recorded in a limitless variety of ways using different instruments, voices, tempos, and interpretations, resulting in an infinite number of valid waveforms representing its musical content.

Conversely, the same waveform may have only a limited number of sheet music representations that can accurately capture its musical elements as per the current tradition and conventions. This is because the unique tonal quality of the waveform is strongly influenced by the specific instrument or production technology used to create it, making it difficult to transcribe its sonic properties into a precise sheet music notation. Moreover, this unique tonal quality is represented by high-level natural language concepts belonging to a rich old cultural and sociological tradition.

>>>>

This thesis argues that high-level musical concepts are invariant to the unique tonal quality of the waveform, much like a sheet music composition can be performed in countless ways using various instruments and styles while retaining its essence.

High-level features like melody, harmony, rhythm, tempo, form, and expression are more abstract and can be recognized regardless of waveform or production style. This approach provides an objective and comprehensive perspective on a piece's musical content and meaning, similar to how reading sheet music reveals a piece's underlying structure and intention.

While a waveform may have a limited number of sheet music representations that accurately capture its musical elements, one sheet music composition can be performed in infinite ways using different instruments, voices, tempos, and interpretations. The unique tonal quality of a waveform is strongly influenced by the specific instrument or production technology used to create it, making it difficult to transcribe its sonic properties into precise sheet music notation.

Additionally, this unique tonal quality is represented by high-level natural language concepts that belong to a rich cultural and sociological tradition.

### 2023-04-07 introduction/Subsections/Introduction.tex
Updates from Overleaf

The reason behind this is non-other but the implementation complexity for such heuristics and abstract fields of study. Closing the gap between human perception, environmental and cultural factors, and computational solutions to mimic the formers is an already tackled topic \cite{Large1994ResonanceMeter}\cite{Mullensiefen2003MeasuringJudgments}\cite{Sears2014PerceivingCadence}\cite{Thaut2014HumanPatter}, yet remains a vast open challenge.

>>>>

The reason behind this is non-other but the implementation complexity for such heuristics and abstract fields of study. Closing the gap between human perception, environmental and cultural factors, and computational solutions to mimic the formers is an already tackled topic \cite{Large1994ResonanceMeter}\cite{Mullensiefen2003MeasuringJudgments}\cite{Sears2014PerceivingCadence}\cite{Thaut2014HumanPatter}\cite{Hernandez-Olivan2023SymbolicMethods}, yet remains a vast open challenge.

### 2023-04-07 introduction/Subsections/Motivation.tex
Updates from Overleaf

Let's pick a core MIR task as a sample to display such complexity: Music Structure Analysis (MSA), an interdisciplinary field that aims to understand the structure of music\cite{Nieto2020Audio-BasedApplications}. However, due to subjectivity, ambiguity, and data scarcity, audio-based MSA faces challenges like boundary placement ambiguity and similarity quantification \cite{NietoPerceptualMusic}. The main principles of MSA were initially defined as homogeneity, novelty, and repetition, with the addition of regularity.

The checkerboard kernel technique is a simple and effective method for Music Structure Analysis (MSA) based on the homogeneity principle. The kernel, with a checkerboard-like structure, is convolved over the main diagonal of a Self-Similarity Matrix (SSM) such as $S_{ij}$ is the similarity between time frames $i$ and $j$, $\vec{x}_i$ is the vector representation of time frame $i$, and $\left| \vec{x}_i \right|$ is the Euclidean norm of vector $\vec{x}_i$. The dot product of two vectors $\vec{x}_i$ and $\vec{x}_j$ is denoted by $\vec{x}_i \cdot \vec{x}_j$.

@@

This yields a novelty curve that highlights sudden changes in the selected musical features (some examples would be chroma, MFCCs, beat-synchronous features, onset and offset detection, and harmonic and percussive separation) from which to extract the segment boundaries.

@@

Combining technical tasks, aural skills, and music perception is challenging due to the complexity and subjectivity of musical perception, discrepancies between the features extracted by DSP techniques and the features perceived by humans, and the variability of musical signals and individual differences in musical perception. DSP techniques may not always reflect how humans perceive music and may not capture the subtle variations in timing or harmonic relationships critical to musical perception. Variability in musical signals due to instrumentation, genre, and performance style further complicates the development of DSP techniques that can be generalized to different musical contexts.

@@

I have consistently been inspired by musicians and researchers who strive to uncover musical ground truth within the existing tonal paradigm, such as Heinrich Schenker \cite{}, or by challenging it, as exemplified by George Russell \cite{LydianRussell}. Both approaches aim to identify abstract concepts that reinforce or disrupt the tonal foundation. Their ultimate goal is to advance the tonal landscape, providing musicians with a dependable playground for growth and development.

@@

\subsection{Assumptions}

>>>>

Let's pick a core MIR task as a sample to display such complexity: Music Structure Analysis (MSA), an interdisciplinary field that aims to understand the structure of music \cite{Nieto2020Audio-BasedApplications}. However, due to subjectivity, ambiguity, and data scarcity, audio-based MSA faces challenges like boundary placement ambiguity and similarity quantification \cite{NietoPerceptualMusic}.

The main principles of MSA were initially defined as homogeneity, novelty, and repetition, with the addition of regularity. The checkerboard kernel technique is a simple and effective method for MSA based on the homogeneity principle. The kernel, with a checkerboard-like structure, is convolved over the main diagonal of a Self-Similarity Matrix (SSM) such as $S_{ij}$ is the similarity between time frames $i$ and $j$, $\vec{x}_i$ is the vector representation of time frame $i$, and $\left| \vec{x}_i \right|$ is the Euclidean norm of vector $\vec{x}_i$. The dot product of two vectors $\vec{x}_i$ and $\vec{x}_j$ is denoted by $\vec{x}_i \cdot \vec{x}_j$.

@@

This yields a novelty curve highlighting sudden changes in the selected musical features from which to extract the segment boundaries.

@@

Combining technical tasks, aural skills, and music perception is challenging due to the complexity and subjectivity of musical perception, discrepancies between the features extracted by DSP techniques and the features perceived by humans, and the variability of musical signals and individual differences in musical perception.

Purely mathematical approaches may not always reflect how humans perceive music and may not capture the subtle variations in timing or harmonic relationships critical to musical perception. Variability in musical signals due to instrumentation, genre, and performance style further complicates the development of these techniques that can be generalized to different musical contexts.

@@

I have consistently been inspired by musicians and researchers who strive to uncover musical ground truth within the existing tonal paradigm, such as Heinrich Schenker \cite{}, or by challenging it, as exemplified by George Russell \cite{LydianRussell}. Both approaches aim to identify abstract concepts that reinforce or disrupt the tonal foundation. Their ultimate goal is to advance the tonal landscape, providing musicians with a dependable playground for growth, development, and further understanding of tonal paradigms.

@@

### 2023-04-07 introduction/Subsections/Objectives.tex
Updates from Overleaf

A self-supervised model that learns an embedding space to retrieve musical content from raw audio signals could have several potential applications and downstream tasks in Music Information Retrieval (MIR). Here are some examples:

>>>>

A self-supervised model that learns an embedding space to retrieve musical content from raw audio signals could have several potential applications and downstream tasks in MIR. Here are some examples:

### 2023-04-07 methods/subsections/model architecture and training strategy.tex
Updates from Overleaf

\section{Model architecture and training strategy}
Piaget's theory of cognitive development \cite{Piagets_Theory_of_Cognitive_Development} posits that children acquire knowledge from sensory and motor experiences in their early stages, from birth to around 18 months. This period, referred to as the \textit{sensorimotor stage}, sees the emergence of early representational thought through basic actions such as sucking, grasping, looking, and listening. As children progress through different developmental stages until adolescence and adulthood, their reasoning progressively moves toward abstract ideas and deductive logic. Schemas, higher-order cognitive structures hypothesized to underlie many aspects of human knowledge and skill, emerge during this process, which Piaget interprets through an equilibration mechanism that balances new information with old knowledge. This mechanism involves \textit{assimilation} and \textit{accommodation}, which entail taking in new information to fit pre-existing schemas and modifying existing schemas resulting from new information. In the context of learning, complex structures are based on simpler ones, and transfer of learning occurs when dynamic structures capable of generalization are created. \cite{audioselfsupsurvey}

>>>>

\section{Model architecture and training strategy}
Piaget's theory of cognitive development \cite{Piagets_Theory_of_Cognitive_Development} posits that children acquire knowledge from sensory and motor experiences in their early stages, from birth to around 18 months. This period referred to as the \textit{sensorimotor stage}, sees the emergence of early representational thought through basic actions such as sucking, grasping, looking, and listening. As children progress through different developmental stages until adolescence and adulthood, their reasoning progressively moves toward abstract ideas and deductive logic. Schemas, higher-order cognitive structures hypothesized to underlie many aspects of human knowledge and skill, emerge during this process, which Piaget interprets through an equilibration mechanism that balances new information with old knowledge. This mechanism involves \textit{assimilation} and \textit{accommodation}, which entail taking in new information to fit pre-existing schemas and modifying existing schemas resulting from new information. In the context of learning, complex structures are based on simpler ones, and transfer of learning occurs when dynamic structures capable of generalization are created. \cite{audioselfsupsurvey}
